---
title: "Heat Vulnerability Analysis in Johannesburg"
output: html_notebook
---

# Introduction

This R Notebook provides a comprehensive analysis of heat vulnerability in Johannesburg using various spatial statistical methods. The analysis includes data preparation, exploratory data analysis, Global and Local Moran's I, Geographically Weighted Principal Component Analysis (GWPCA), and Geographically Weighted Regression (GWR). The notebook incorporates corrected code and explanations to guide you through each step.

# Load Required Libraries

First, we need to load all the necessary libraries for the analysis. If any of these packages are not installed on your system, you can install them using `install.packages("package_name")`.

```{r}
# Load necessary libraries
library(sf)            # For handling spatial data
library(sp)            # Spatial data classes
library(dplyr)         # Data manipulation
library(ggplot2)       # Data visualization
library(corrplot)      # Correlation plots
library(psych)         # PCA and psychometric analysis
library(GWmodel)       # Geographically Weighted Models
library(spdep)         # Spatial dependence analysis
library(factoextra)    # Visualization for PCA
library(spgwr)         # Geographically Weighted Regression
library(spData)        # Sample spatial data
library(RColorBrewer)  # Color palettes
library(viridis)       # Color palettes
```
```{r}
# Set working directory
setwd("C:/Users/CraigParker/OneDrive - Wits PHR/Desktop/HVI_Johannesburg")
```

# Data Preparation

In this section, we will read the spatial data and perform initial data preparation, including handling missing values and ensuring variables are correctly formatted.
```{r}
# Read data
data <- read.csv('data.csv', header = TRUE)
geometry <- st_read("geometry.shp")
geometry <- st_transform(geometry, crs = 32735)
```


```{r}
# Load necessary libraries
library(sf)            # For handling spatial data
library(sp)            # Spatial data classes
library(dplyr)         # Data manipulation
library(ggplot2)       # Data visualization
library(corrplot)      # Correlation plots
library(psych)         # PCA and psychometric analysis
library(GWmodel)       # Geographically Weighted Models
library(spdep)         # Spatial dependence analysis
library(factoextra)    # Visualization for PCA
library(spgwr)         # Geographically Weighted Regression
library(spData)        # Sample spatial data
library(RColorBrewer)  # Color palettes
library(viridis)       # Color palettes
library(tidyr)         # Data reshaping (for pivot_longer)

# Set working directory
setwd("C:/Users/CraigParker/OneDrive - Wits PHR/Desktop/HVI_Johannesburg")

# Read data
data <- read.csv('data.csv', header = TRUE)
geometry <- st_read("geometry.shp")
geometry <- st_transform(geometry, crs = 32735)

# Ensure consistent IDs (replace 'WardID_' with your actual key column)
data$WardID_ <- as.character(data$WardID_)
geometry$WardID_ <- as.character(geometry$WardID_)

# Merge datasets
merged_data <- merge(
  geometry[, c("WardID_", "geometry")],
  data,
  by = "WardID_",
  all.x = TRUE
)

# Define variables
vars <- c(
  "Crowded.dwellings", "No.piped.water", "Using.public.healthcare.facilities",
  "Poor.health.status", "Failed.to.find.healthcare.when.needed",
  "No.medical.insurance", "Household.hunger.risk",
  "Benefiting.from.school.feeding.scheme", "UTFVI", "LST", "NDVI",
  "NDBI__mean", "concern_he", "cancer_pro", "diabetes_p", "pneumonia_",
  "heart_dise", "hypertensi", "hiv_prop", "tb_prop", "covid_prop",
  "X60_plus_pr"
)

# Check for missing variables
missing_vars <- setdiff(vars, names(merged_data))
if (length(missing_vars) > 0) {
  stop("Missing variables: ", paste(missing_vars, collapse = ", "))
}

# Ensure variables are numeric
for (var in vars) {
  if (!is.numeric(merged_data[[var]])) {
    merged_data[[var]] <- as.numeric(as.character(merged_data[[var]]))
  }
}

# Remove rows with missing data
merged_data_no_geom <- st_drop_geometry(merged_data)
rows_complete <- complete.cases(merged_data_no_geom[vars])
merged_data <- merged_data[rows_complete, ]
merged_data_no_geom <- merged_data_no_geom[rows_complete, ]

# Compute descriptive statistics
summary_stats <- merged_data_no_geom %>%
  select(all_of(vars)) %>%
  summarise(across(everything(), list(
    mean = ~mean(.),
    sd = ~sd(.),
    median = ~median(.),
    min = ~min(.),
    max = ~max(.)
  )))

# Reshape to long format using names_pattern
summary_stats_long <- summary_stats %>%
  pivot_longer(
    cols = everything(),
    names_to = c("Variable", "Statistic"),
    names_pattern = "^(.*)_(mean|sd|median|min|max)$"
  )

# View the summary statistics
print(summary_stats_long)

# Compute correlation matrix
corr_matrix <- cor(
  merged_data_no_geom %>% select(all_of(vars)),
  use = "complete.obs"
)

# Visualize correlation matrix
corrplot(corr_matrix, method = "circle")






```

# Exploratory Data Analysis

Let's compute and display summary statistics for the selected variables.

```{r}
# Compute summary statistics
summary_stats <- gwpca_sf %>%
  st_set_geometry(NULL) %>%
  summarise(across(all_of(vars), list(mean = mean, sd = sd, median = median, min = min, max = max)))

# Reshape data for better display
library(tidyr)
summary_stats_long <- summary_stats %>%
  pivot_longer(cols = everything(),
               names_to = c("Variable", "Statistic"),
               names_sep = "_",
               values_to = "Value")

print(summary_stats_long)
```

# Correlation Matrix and Heatmap

Compute the correlation matrix and visualize it using a heatmap.

```{r}
# Compute correlation matrix
corr_matrix <- cor(gwpca_sf %>% st_set_geometry(NULL) %>% select(all_of(vars)), use = "complete.obs")

# Plot correlation heatmap
corrplot(corr_matrix, method = "color", type = "upper", order = "hclust",
         tl.cex = 0.7, tl.col = "black", tl.srt = 45,
         addCoef.col = "black", number.cex = 0.7,
         title = "Correlation Matrix of Variables")
```

# Principal Component Analysis (PCA)

Perform PCA to reduce dimensionality and identify key components contributing to heat vulnerability.

```{r}
# Perform PCA
pca_result <- prcomp(gwpca_sf %>% st_set_geometry(NULL) %>% select(all_of(vars)), scale. = TRUE)

# Visualize PCA variable loadings
fviz_pca_var(pca_result,
             col.var = "contrib", # Color by contributions to the PCs
             gradient.cols = c("blue", "yellow", "red"),
             repel = TRUE) +
  labs(title = "PCA Variable Loadings")
```

# Calculation of Heat Vulnerability Index (HVI)

Compute the HVI using the first principal component.

```{r}
# Extract the first principal component
gwpca_sf$HVI <- pca_result$x[,1]

# Normalize HVI to a 0-1 scale
gwpca_sf$HVI <- scales::rescale(gwpca_sf$HVI, to = c(0, 1))
```

# Mapping the Heat Vulnerability Index

Visualize the spatial distribution of HVI across Johannesburg.

```{r}
# Plot HVI map
plot_hvi <- ggplot(gwpca_sf) +
  geom_sf(aes(fill = HVI)) +
  scale_fill_viridis_c(option = "inferno") +
  theme_minimal() +
  labs(title = "Heat Vulnerability Index", fill = "HVI")

print(plot_hvi)
```

# Identifying Top 10 Most Vulnerable Wards

Highlight the top 10 wards with the highest HVI.

```{r}
# Select top 10 most vulnerable wards
top_10_vulnerable_wards <- gwpca_sf %>%
  arrange(desc(HVI)) %>%
  slice(1:10)

# Plot the top 10 wards
plot_top_10_hvi <- ggplot() +
  geom_sf(data = gwpca_sf, aes(fill = HVI), alpha = 0.6) +
  geom_sf(data = top_10_vulnerable_wards, fill = "black", color = "black", size = 0.7, alpha = 0.8) +
  scale_fill_viridis_c(option = "inferno") +
  theme_minimal() +
  labs(title = "Top 10 Most Vulnerable Wards in Johannesburg", fill = "HVI")

print(plot_top_10_hvi)
```

# Global and Local Moran's I

Assess spatial autocorrelation in HVI using Global and Local Moran's I statistics.

```{r}
library(spdep)

# Convert sf object to spatial object for spdep
gwpca_sp <- as(gwpca_sf, "Spatial")

# Create spatial weights matrix
neighbors <- poly2nb(gwpca_sf)
weights <- nb2listw(neighbors, style = "W")

# Calculate Global Moran's I
global_moran <- moran.test(gwpca_sf$HVI, listw = weights)
print(global_moran)

# Calculate Local Moran's I
local_moran <- localmoran(gwpca_sf$HVI, listw = weights)

# Add Local Moran's I results to the sf dataframe
gwpca_sf$LocalI <- local_moran[, "Ii"]
gwpca_sf$LocalI_pvalue <- local_moran[, "Pr(z != E(Ii))"]

# Map the Local Moran's I
plot_lisa <- ggplot(gwpca_sf) +
  geom_sf(aes(fill = LocalI)) +
  scale_fill_distiller(palette = "RdBu", direction = 1) +
  theme_minimal() +
  labs(title = "Local Moran's I for HVI", fill = "Local I")

print(plot_lisa)
```

# Geographically Weighted Principal Component Analysis (GWPCA)

Perform GWPCA to understand how the principal components vary spatially.

```{r}
library(GWmodel)

# Define bandwidth for GWPCA
bw_gwpca <- bw.gwpca(gwpca_sf %>% st_set_geometry(NULL) %>% select(all_of(vars)),
                     coords = st_coordinates(st_centroid(gwpca_sf)),
                     k = 20)

# Perform GWPCA
gwpca_result <- gwpca(gwpca_sf %>% st_set_geometry(NULL) %>% select(all_of(vars)),
                      coords = st_coordinates(st_centroid(gwpca_sf)),
                      bw = bw_gwpca,
                      k = 2)

# Extract local scores and add to spatial data
local_scores <- as.data.frame(gwpca_result$loadings)
gwpca_sf$LocalPC1 <- local_scores[,1]
gwpca_sf$LocalPC2 <- local_scores[,2]
```

# Mapping Local Principal Components

Visualize the spatial variation of the first local principal component.

```{r}
# Map Local PC1
plot_local_pc1 <- ggplot(gwpca_sf) +
  geom_sf(aes(fill = LocalPC1)) +
  scale_fill_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(title = "Local PC1 from GWPCA", fill = "Local PC1")

print(plot_local_pc1)
```

# Geographically Weighted Regression (GWR)

Conduct GWR to model the relationship between HVI and contributing variables, allowing coefficients to vary spatially.

```{r}
library(spgwr)

# Define dependent and independent variables
dependent_var <- "HVI"
independent_vars <- c(
  "UTFVI", "NDBI_mean", "LST", "NDVI", "No.medical.insurance",
  "Using.public.healthcare.facilities", "Poor.health.status",
  "hypertensi", "X60_plus_pr"
)

# Ensure variables are present and correctly named
missing_vars <- independent_vars[!independent_vars %in% names(gwpca_sf)]
if (length(missing_vars) > 0) {
  print("The following variables are missing from gwpca_sf:")
  print(missing_vars)
}

# Remove missing variables from the list
independent_vars <- independent_vars[independent_vars %in% names(gwpca_sf)]

# Ensure variables are numeric and scaled
for (var in independent_vars) {
  gwpca_sf[[var]] <- as.numeric(gwpca_sf[[var]])
  gwpca_sf[[var]] <- scale(gwpca_sf[[var]])
}

# Remove missing values
gwpca_sf_complete <- gwpca_sf %>% 
  drop_na(all_of(c(dependent_var, independent_vars)))

# Coordinates for GWR
coords <- st_coordinates(st_centroid(gwpca_sf_complete))

# Bandwidth selection
gwr_bw <- gwr.sel(
  formula = as.formula(paste(dependent_var, "~", paste(independent_vars, collapse = "+"))),
  data = st_set_geometry(gwpca_sf_complete, NULL),
  coords = coords,
  adapt = TRUE
)

print(paste("Optimal bandwidth for GWR:", gwr_bw))
```

# Running the GWR Model

```{r}
# Run GWR model
gwr_model <- gwr(
  formula = as.formula(paste(dependent_var, "~", paste(independent_vars, collapse = "+"))),
  data = st_set_geometry(gwpca_sf_complete, NULL),
  coords = coords,
  adapt = gwr_bw,
  hatmatrix = TRUE
)

# Extract results
gwr_results <- as.data.frame(gwr_model$SDF)

# Add results to spatial dataframe
gwpca_sf_complete <- cbind(gwpca_sf_complete, gwr_results)
```

# Mapping GWR Coefficients

Visualize the spatial variation of the coefficient for one of the key variables, e.g., LST.

```{r}
# Map LST coefficient
plot_gwr_lst <- ggplot(gwpca_sf_complete) +
  geom_sf(aes(fill = LST)) +
  scale_fill_distiller(palette = "RdBu", direction = 1) +
  theme_minimal() +
  labs(title = "GWR Coefficient for LST", fill = "LST Coefficient")

print(plot_gwr_lst)
```

# Conclusion

This notebook provided a step-by-step analysis of heat vulnerability in Johannesburg, incorporating corrected code and explanations. We performed data preparation, exploratory data analysis, PCA, GWPCA, and GWR, and visualized the results to gain insights into the spatial dynamics of heat vulnerability.

# Notes

- **Data Paths:** Replace `"path/to/your/geometry.shp"` with the actual file path to your spatial data.
- **Variable Names:** Ensure that the variable names used in the code match exactly with those in your data.
- **Package Installation:** If any packages are not installed, you can install them using `install.packages("package_name")`.
- **CRS Transformation:** Ensure that all spatial data are in the same Coordinate Reference System (CRS) for accurate spatial analysis.

# References

- **GWmodel Package:** [GWmodel Documentation](https://cran.r-project.org/web/packages/GWmodel/GWmodel.pdf)
- **spdep Package:** [spdep Documentation](https://cran.r-project.org/web/packages/spdep/spdep.pdf)
- **factoextra Package:** [factoextra Documentation](https://cran.r-project.org/web/packages/factoextra/factoextra.pdf)

# Session Information

```{r}
sessionInfo()
```