I'm sorry to hear that you lost your notebook. Based on our previous conversation, I've reconstructed the entire notebook for your heat vulnerability analysis in Johannesburg. This notebook includes all the corrected code, enhanced visualizations, and explanations to guide you through each step of the analysis.

---

# Heat Vulnerability Analysis in Johannesburg

This R Notebook provides a comprehensive analysis of heat vulnerability in Johannesburg using various spatial statistical methods. The analysis includes:

- **Data Preparation**
- **Exploratory Data Analysis**
- **Correlation Matrix Visualization with Grouped Variables**
- **Principal Component Analysis (PCA) with Enhanced Visualization**
- **Calculation and Mapping of the Heat Vulnerability Index (HVI)**
- **Spatial Autocorrelation Analysis (Global and Local Moran's I)**
- **Geographically Weighted Principal Component Analysis (GWPCA)**

---

```{r setup, include=FALSE}
# Set global options
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Load Required Libraries

First, we need to load all the necessary libraries for the analysis.

```{r}
# Load necessary libraries
library(sf)            # For handling spatial data
library(dplyr)         # Data manipulation
library(ggplot2)       # Data visualization
library(corrplot)      # Correlation plots
library(psych)         # PCA and psychometric analysis
library(GWmodel)       # Geographically Weighted Models
library(spdep)         # Spatial dependence analysis
library(factoextra)    # Visualization for PCA
library(RColorBrewer)  # Color palettes
library(viridis)       # Color palettes
library(tidyr)         # Data reshaping
library(scales)        # Scaling data
library(ggcorrplot)    # Enhanced correlation plots
library(plotly)        # Interactive plots
```

## Set Working Directory

Set your working directory to the location where your data files are stored.

```{r}
# Set working directory
setwd("C:/Users/CraigParker/OneDrive - Wits PHR/Desktop/HVI_Johannesburg")
```

## Data Preparation

Read the spatial data and perform initial data preparation, including handling missing values and ensuring variables are correctly formatted.

```{r}
# Read spatial data
gwpca_sf <- st_read("C:/Users/CraigParker/OneDrive - Wits PHR/Desktop/HVI_Johannesburg", layer = "geometry")

# If the data is in a different CRS, transform it to a common CRS
gwpca_sf <- st_transform(gwpca_sf, crs = 32735)  # WGS 84 / UTM zone 35S
```

### Variable Selection

Ensure that the variable names used in the code match exactly with those in your data.

```{r}
# List of variables to be used in the analysis (ensure these match exactly)
vars <- c(
  "Crowded.dw",     # Crowded dwellings
  "No.piped.w",     # No piped water
  "Using.publ",     # Using public healthcare facilities
  "Poor.healt",     # Poor health status
  "Failed.to",      # Failed to find healthcare when needed
  "No.medical",     # No medical insurance
  "Household",      # Household hunger risk
  "Benefiting",     # Benefiting from school feeding scheme
  "UTFVI",          # Urban Thermal Field Variance Index
  "LST",            # Land Surface Temperature
  "NDVI",           # Normalized Difference Vegetation Index
  "NDBI__mean",     # Normalized Difference Built-up Index
  "concern_he", 
  "cancer_pro", 
  "diabetes_p", 
  "pneumonia_", 
  "heart_dise", 
  "hypertensi", 
  "hiv_prop", 
  "tb_prop", 
  "covid_prop", 
  "X60_plus_pr"     # Population aged 60+
)
```

### Verify Variable Names

```{r}
# Inspect column names of the spatial data
names(gwpca_sf)

# Identify variables in 'vars' that are not in 'gwpca_sf'
missing_vars <- vars[!vars %in% names(gwpca_sf)]
if(length(missing_vars) > 0){
  cat("The following variables are missing from gwpca_sf:\n")
  print(missing_vars)
} else {
  cat("All variables in 'vars' are present in 'gwpca_sf'.\n")
}
```

### Subset Data

```{r}
# Subset the data to include only the variables in 'vars' and geometry
gwpca_data <- gwpca_sf %>%
  select(all_of(vars), geometry)

# Print the first few rows to verify the selected data
head(gwpca_data)
```

### Ensure Variables are Numeric

```{r}
# Convert variables to numeric
for (var in vars) {
  gwpca_data[[var]] <- as.numeric(gwpca_data[[var]])
}
```

### Handle Missing Values

```{r}
# Remove rows with missing values in the selected variables
gwpca_data <- gwpca_data %>%
  drop_na(all_of(vars))
```

## Exploratory Data Analysis

### Summary Statistics

Compute and display summary statistics for the selected variables.

```{r}
# Compute summary statistics
summary_stats <- gwpca_data %>%
  st_set_geometry(NULL) %>%
  summarise(across(everything(), list(
    mean = mean, sd = sd, median = median, min = min, max = max
  )))

# Reshape data for better display
summary_stats_long <- summary_stats %>%
  pivot_longer(cols = everything(),
               names_to = c("Variable", "Statistic"),
               names_sep = "_",
               values_to = "Value")

print(summary_stats_long)
```

### Correlation Matrix and Enhanced Visualization

#### Define Variable Groups and Labels

```{r}
# Define variable groups
variable_groups <- c(
  "Crowded.dw" = "Socioeconomic",
  "No.piped.w" = "Socioeconomic",
  "Using.publ" = "Socioeconomic",
  "Poor.healt" = "Health",
  "Failed.to" = "Health",
  "No.medical" = "Socioeconomic",
  "Household" = "Socioeconomic",
  "Benefiting" = "Socioeconomic",
  "UTFVI" = "Environmental",
  "LST" = "Environmental",
  "NDVI" = "Environmental",
  "NDBI__mean" = "Environmental",
  "concern_he" = "Health",
  "cancer_pro" = "Health",
  "diabetes_p" = "Health",
  "pneumonia_" = "Health",
  "heart_dise" = "Health",
  "hypertensi" = "Health",
  "hiv_prop" = "Health",
  "tb_prop" = "Health",
  "covid_prop" = "Health",
  "X60_plus_pr" = "Socioeconomic"
)

# Update variable labels to be more descriptive
variable_labels <- c(
  "Crowded.dw" = "Crowded Dwellings",
  "No.piped.w" = "No Piped Water",
  "Using.publ" = "Public Healthcare Use",
  "Poor.healt" = "Poor Health Status",
  "Failed.to" = "Failed to Find Healthcare",
  "No.medical" = "No Medical Insurance",
  "Household" = "Household Hunger Risk",
  "Benefiting" = "School Feeding Scheme",
  "UTFVI" = "UTFVI",
  "LST" = "Land Surface Temp",
  "NDVI" = "NDVI",
  "NDBI__mean" = "NDBI Mean",
  "concern_he" = "Health Concern",
  "cancer_pro" = "Cancer Proportion",
  "diabetes_p" = "Diabetes Proportion",
  "pneumonia_" = "Pneumonia Proportion",
  "heart_dise" = "Heart Disease Proportion",
  "hypertensi" = "Hypertension Proportion",
  "hiv_prop" = "HIV Proportion",
  "tb_prop" = "TB Proportion",
  "covid_prop" = "COVID Proportion",
  "X60_plus_pr" = "60+ Population Proportion"
)
```

#### Create Variable Information Data Frame

```{r}
# Create a data frame with variable names, groups, and labels
variable_info <- data.frame(
  variable = names(variable_groups),
  group = unname(variable_groups),
  label = unname(variable_labels),
  stringsAsFactors = FALSE
)
```

#### Compute Correlation Matrix

```{r}
# Compute correlation matrix
corr_matrix <- gwpca_data %>%
  st_set_geometry(NULL) %>%
  cor(use = "complete.obs")
```

#### Reorder Variables and Correlation Matrix

```{r}
# Order variables by group
variable_info_ordered <- variable_info %>%
  arrange(factor(group, levels = c("Environmental", "Health", "Socioeconomic")), variable)

ordered_variables <- variable_info_ordered$variable
variable_labels_ordered <- variable_info_ordered$label

# Reorder the correlation matrix
corr_matrix_ordered <- corr_matrix[ordered_variables, ordered_variables]

# Set the row and column names to the descriptive labels
rownames(corr_matrix_ordered) <- variable_labels_ordered
colnames(corr_matrix_ordered) <- variable_labels_ordered
```

#### Assign Colors to Groups

```{r}
# Assign colors to groups
group_colors <- c("Environmental" = "forestgreen", "Health" = "darkorange", "Socioeconomic" = "purple")
label_colors_ordered <- group_colors[variable_info_ordered$group]
```

#### Plot the Enhanced Correlation Matrix

```{r}
# Define color palette with red for positive correlations and blue for negative correlations
corr_colors <- colorRampPalette(c("blue", "white", "red"))(200)

# Adjust label size and rotation
label_cex <- 0.6  # Adjust this value to change label size
label_srt <- 45   # Adjust this value to change label rotation

# Plot the correlation matrix with updated labels and color scheme
corrplot(corr_matrix_ordered,
         method = "color",
         col = corr_colors,
         type = "upper",
         order = "original",
         tl.cex = label_cex,
         tl.col = label_colors_ordered,
         tl.srt = label_srt,
         addCoef.col = "black",
         number.cex = 0.5,
         mar = c(0,0,2,0))

# Add a legend for variable groups
legend("topright", legend = names(group_colors), col = group_colors, pch = 15, cex = 0.8)
```

**Note:** Adjust `label_cex` and `label_srt` to fit the labels appropriately.

## Principal Component Analysis (PCA) with Enhanced Visualization

Perform PCA to reduce dimensionality and identify key components contributing to heat vulnerability.

### Perform PCA

```{r}
# Prepare data for PCA
pca_data <- gwpca_data %>%
  st_set_geometry(NULL) %>%
  drop_na()

# Perform PCA
pca_result <- prcomp(pca_data, scale. = TRUE)
```

### Define Variable Groups (if not already defined)

```{r}
# Variable groups and labels are already defined in previous steps
```

### Enhance the PCA Variable Loadings Plot

```{r}
# Load necessary libraries
library(factoextra)
library(ggplot2)

# Create a vector of colors for the groups
group_colors <- c("Environmental" = "forestgreen", "Health" = "darkorange", "Socioeconomic" = "purple")

# Match colors to variables
variable_colors <- group_colors[variable_info$group]
names(variable_colors) <- variable_info$variable

# Plot the PCA variable loadings with color-coded groups
fviz_pca_var(pca_result,
             geom = c("arrow", "text"),
             col.var = variable_info$group, # Color by groups
             palette = group_colors,
             legend.title = "Variable Groups",
             repel = TRUE) +
  labs(title = "PCA Variable Loadings Grouped by Category")
```

### Explore Additional Principal Components

```{r}
# Plot PC1 vs PC3
fviz_pca_var(pca_result,
             axes = c(1, 3),
             geom = c("arrow", "text"),
             col.var = variable_info$group,
             palette = group_colors,
             legend.title = "Variable Groups",
             repel = TRUE) +
  labs(title = "PCA Variable Loadings (PC1 vs PC3)")
```

## Calculation of Heat Vulnerability Index (HVI)

Compute the HVI using the first principal component.

```{r}
# Extract the first principal component
gwpca_data$HVI <- pca_result$x[, 1]

# Normalize HVI to a 0-1 scale
gwpca_data$HVI <- rescale(gwpca_data$HVI, to = c(0, 1))
```

## Mapping the Heat Vulnerability Index

Visualize the spatial distribution of HVI across Johannesburg.

```{r}
# Plot HVI map
plot_hvi <- ggplot(gwpca_data) +
  geom_sf(aes(fill = HVI)) +
  scale_fill_viridis_c(option = "inferno") +
  theme_minimal() +
  labs(title = "Heat Vulnerability Index", fill = "HVI")

print(plot_hvi)
```

## Identifying Top 10 Most Vulnerable Wards

Highlight the top 10 wards with the highest HVI.

```{r}
# Select top 10 most vulnerable wards
top_10_vulnerable_wards <- gwpca_data %>%
  arrange(desc(HVI)) %>%
  slice(1:10)

# Plot the top 10 wards
plot_top_10_hvi <- ggplot() +
  geom_sf(data = gwpca_data, aes(fill = HVI), alpha = 0.6) +
  geom_sf(data = top_10_vulnerable_wards, fill = "black", color = "black", size = 0.7, alpha = 0.8) +
  scale_fill_viridis_c(option = "inferno") +
  theme_minimal() +
  labs(title = "Top 10 Most Vulnerable Wards in Johannesburg", fill = "HVI")

print(plot_top_10_hvi)
```

## Global and Local Moran's I

Assess spatial autocorrelation in HVI using Global and Local Moran's I statistics.

```{r}
# Convert sf object to spatial object for spdep
gwpca_sp <- as_Spatial(gwpca_data)

# Create spatial weights matrix
neighbors <- poly2nb(gwpca_sp)
weights <- nb2listw(neighbors, style = "W")

# Calculate Global Moran's I
global_moran <- moran.test(gwpca_data$HVI, listw = weights)
print(global_moran)

# Calculate Local Moran's I
local_moran <- localmoran(gwpca_data$HVI, listw = weights)

# Add Local Moran's I results to the sf dataframe
gwpca_data$LocalI <- local_moran[, "Ii"]
gwpca_data$LocalI_pvalue <- local_moran[, "Pr(z != E(Ii))"]
```

### Mapping Local Moran's I

Visualize the spatial variation of Local Moran's I.

```{r}
# Map the Local Moran's I
plot_lisa <- ggplot(gwpca_data) +
  geom_sf(aes(fill = LocalI)) +
  scale_fill_distiller(palette = "RdBu", direction = 1) +
  theme_minimal() +
  labs(title = "Local Moran's I for HVI", fill = "Local I")

print(plot_lisa)
```

## Geographically Weighted Principal Component Analysis (GWPCA)

Perform GWPCA to understand how the principal components vary spatially.

```{r}
## Step 1: Examine the gwpca_result object
str(gwpca_result)

# Step 2: Verify function arguments
args(gwpca)

# Step 3: Update the function call (if necessary)
gwpca_result <- gwpca(
  data = gwpca_data_sp,
  vars = vars,
  k = 2,
  robust = TRUE,
  bw = bw
  # Removed adaptive and scores arguments if not recognized
)

# Step 4: Try different bandwidth values
bw_smaller <- 50  # Adjust as needed
gwpca_result <- gwpca(
  data = gwpca_data_sp,
  vars = vars,
  k = 2,
  robust = TRUE,
  bw = bw_smaller,
  adaptive = TRUE
)

# Step 5: Use fixed bandwidth
bw_fixed <- 10000  # Adjust as appropriate
gwpca_result <- gwpca(
  data = gwpca_data_sp,
  vars = vars,
  k = 2,
  robust = TRUE,
  bw = bw_fixed,
  adaptive = FALSE
)

# Step 6: Check for data issues (see code above)

# Step 7: Adjust k
gwpca_result <- gwpca(
  data = gwpca_data_sp,
  vars = vars,
  k = 1,
  robust = TRUE,
  bw = bw,
  adaptive = TRUE
)

# Step 8: Check package version
packageVersion("GWmodel")
# Update if necessary

# Step 9: Use the formula interface
pca_formula <- as.formula(paste("~", paste(vars, collapse = "+")))
gwpca_result <- gwpca(
  formula = pca_formula,
  data = gwpca_data_sp,
  k = 2,
  robust = TRUE,
  bw = bw,
  adaptive = TRUE
)


```
```{r}
# Print the gwpca_result object
print(gwpca_result)

```

### Mapping Local Principal Components
```{r}
## Geographically Weighted Principal Component Analysis (GWPCA)

# Load GWmodel package
library(GWmodel)

# Prepare data for GWPCA
gwpca_data_complete <- gwpca_data

# Convert sf to Spatial*DataFrame for GWmodel
gwpca_data_sp <- as(gwpca_data_complete, "Spatial")

# Manually specify a smaller bandwidth (e.g., 50)
bw_manual <- 50

# Perform GWPCA with pca.gw = TRUE and specified bandwidth
gwpca_result <- gwpca(
  data = gwpca_data_sp,
  vars = vars,
  k = 2,
  robust = TRUE,
  bw = bw_manual,
  adaptive = TRUE,
  pca.gw = TRUE,       # Enable local PCA
  scores = TRUE,
  verbose = TRUE       # Display messages during execution
)

# Check the gwpca_result object
print(gwpca_result)

# Extract local scores and add to spatial data
if (!is.null(gwpca_result$local.scores)) {
  local_scores <- as.data.frame(gwpca_result$local.scores)
  
  # Verify that local_scores has the expected dimensions
  print(dim(local_scores))
  
  # Ensure that local_scores has at least two columns
  if (ncol(local_scores) >= 2) {
    gwpca_data_complete$LocalPC1 <- local_scores[, 1]
    gwpca_data_complete$LocalPC2 <- local_scores[, 2]
  } else {
    stop("Local scores do not have enough components.")
  }
} else {
  stop("Local scores are not available in gwpca_result.")
}

```

Visualize the spatial variation of the first local principal component.

```{r}
# Map Local PC1
plot_local_pc1 <- ggplot(gwpca_data_complete) +
  geom_sf(aes(fill = LocalPC1)) +
  scale_fill_viridis_c(option = "plasma") +
  theme_minimal() +
  labs(title = "Local PC1 from GWPCA", fill = "Local PC1")

print(plot_local_pc1)
```

## Conclusion

This notebook provided a comprehensive analysis of heat vulnerability in Johannesburg, incorporating corrected code and enhanced visualizations. We performed data preparation, exploratory data analysis, PCA with enhanced visualization, calculation and mapping of HVI, spatial autocorrelation analysis, and GWPCA to gain insights into the spatial dynamics of heat vulnerability.

## Notes

- **Data Paths:** Ensure that the file paths to your spatial data are correct.
- **Variable Names:** We have updated the variable names to match exactly with those in your data.
- **Package Installation:** If any packages are not installed, you can install them using `install.packages("package_name")`.
- **CRS Transformation:** Ensure that all spatial data are in the same Coordinate Reference System (CRS) for accurate spatial analysis.
- **Handling Missing Data:** We used `drop_na()` to remove rows with missing data for analyses that require complete cases.
- **Numeric Variables:** All variables used in PCA and GWPCA must be numeric.
- **Visualizations:** We enhanced the correlation matrix and PCA plots by grouping variables and adjusting color schemes for better interpretation.

## References

- **GWmodel Package:** [GWmodel Documentation](https://cran.r-project.org/web/packages/GWmodel/GWmodel.pdf)
- **spdep Package:** [spdep Documentation](https://cran.r-project.org/web/packages/spdep/spdep.pdf)
- **factoextra Package:** [factoextra Documentation](https://cran.r-project.org/web/packages/factoextra/factoextra.pdf)
- **corrplot Package:** [corrplot Documentation](https://cran.r-project.org/web/packages/corrplot/corrplot.pdf)
- **ggcorrplot Package:** [ggcorrplot Documentation](https://cran.r-project.org/web/packages/ggcorrplot/ggcorrplot.pdf)

## Session Information

```{r}
sessionInfo()
```

---

**Please note**: The code provided assumes that all necessary data files are correctly placed and that the variable names match those in your datasets. Adjust paths and variable names as needed based on your actual data.

Let me know if you have any questions or need further assistance!